Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 appendix.tex
going through re-structured paper

In the vertically global shearing box (VGSB) approximation, we set
$\hat{g}_c=0$ in Eq. \ref{lin_mass} --- \ref{lin_energy}. We then eliminate
the horizonatal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{align}
  & \ii\zeta\delta v_z = \frac{dW}{dz} + \frac{d\ln{\rho}}{d z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\zeta^2}{c_s^2}Q + \frac{\zeta^2k_x^2}{D}W =\ii\zeta \left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\delta v_z - \ii\zeta \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\zeta^2W - \gamma\zeta^2 Q +
  \frac{\ii\zeta}{t_c}\left(W-Q\right) 
  =\ii \zeta c_s^2\frac{d\ln{\rho}}{dz}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}
(note that $\Gamma=1$ has been used) where
\begin{align}
  D \equiv \kappa^2 - \zeta^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

@@

\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\zeta^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\zeta t_c\right)/\left(1-\ii\zeta
t_c \gamma\right)$. Making the low frequency and Keplerian
approximations is equivalent to the replacement $D\to
\Omega_K^2$. This gives, in terms of dimensionless variables,  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\zeta}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align} 
Since we are interested in thin disks with $h \ll 1$ and $|q| =
O(1)$, we neglect the terms proportional to $h^2$, giving 
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\zeta}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
 \end{align}
 which is used in the discussion below.  Eq. \ref{nearly_iso_explicit}, as
 used in our analytical discussion, is obtained by inserting the 
 thin-disk expression for the density field. We comment that the simpler way
 of obtaining Eq. \ref{vertiso_gov_nondim} is to neglect the vertical 
 dependence of $D$ by setting $D\to\Omega_K^2$ in Eq. \ref{ode_w} before eliminating
 variables in favor of $\delta v_z$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\zeta}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\zeta}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\zeta}^2$ is real, and  $\hat{\zeta}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\zeta}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\zeta}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii h q \hat{k} f(\zhat)
      }{1-\hat{\zeta}^2} \right]W^\prime +
  \hat{\zeta}^2\left(1+\frac{\hat{k}^2}{1-\hat{\zeta}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = h^2q f(\hat{z})\Omega_K^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+h^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

\subsection{Stability conditions in the low-frequency limit}
In the low-frequency limit we assume $|\hat{\zeta}|\ll 1$ and
approximate Eq. \ref{iso_ode0} as 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii h q \hat{k}
    f(\zhat)\right]W^\prime +
  \hat{\zeta}^2\left(1+\hat{k}^2\right)W. \label{iso_ode1}
\end{align}
\subsubsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  \hat{\zeta}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat
  =\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii h q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\zeta}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. there must
be vertical shear. A similar conclusion can be reached in the
equivalent linear problem in global cylindrical disk geometry.

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\zeta} = \hat{\omega} +
% \ii\hat{\zeta}$ with real $\hat{\omega}$ and $\hat{\zeta}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\sigma}^2\right)\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\sigma}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\zeta}$ and
$\hat{\sigma}=\imag\hat{\zeta}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\zeta}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\sigma}|\leq
  \frac{h |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|) < \frac{h |q|}{2}\mathrm{max}(|f|) . 
\end{align}
%Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), 
It
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered,
\begin{align}
  \sigma < \mathrm{max} \left|r\frac{d\Omega}{dz}\right|, 
\end{align}
as expected on physical grounds.

@@

\subsection{Explicit solutions in the thin-disk limit}\label{iso_explicit}
Here we assume a thin disk ($h\ll1$) so that $\ln\rho\simeq
-\zhat^2/2$ and $f(\zhat)\simeq \zhat$. However, we do not assume 
low frequency from the onset. Then Eq. \ref{iso_ode0} becomes 
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \frac{\ii qh \hat{k}}{1-\hat{\zeta}^2}\right)\hat{z}W^\prime  +
  \hat{\zeta}^2\left(1+\frac{\hat{k}^2}{1-\hat{\zeta}^2}\right)W = 0.
\end{align}
We remark that taking the low frequency limit and considering
$\hat{k}^2\gg 1$, Eq. \ref{iso_ode3} becomes equivalent to Eq. 39 in
\citetalias{nelson13} or Eq. 28 in \citetalias{barker15}, although we have taken a
different route.

>>>>

In the vertically global shearing box (VGSB) approximation, we set
$\hat{g}_c=0$ in Eq. \ref{lin_mass} --- \ref{lin_energy}. We then eliminate
the horizonatal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{d\ln{\rho}}{d z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{d\ln{\rho}}{dz}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}
(note that $\Gamma=1$ has been used) where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

@@

\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\upsilon t_c\right)/\left(1-\ii\upsilon
t_c \gamma\right)$. Making the low frequency and Keplerian
approximations is equivalent to the replacement $D\to
\Omega_K^2$. This gives, in terms of dimensionless variables,  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align} 
Since we are interested in thin disks with $h \ll 1$ and $|q| =
O(1)$, we neglect the terms proportional to $h^2$, giving 
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
 \end{align}
 which is used in the discussion below.  Eq. \ref{nearly_iso_explicit}, as
 used in our analytical discussion, is obtained by inserting the 
 thin-disk expression for the density field. We comment that the simpler way
 of obtaining Eq. \ref{vertiso_gov_nondim} is to neglect the vertical 
 dependence of $D$ by setting $D\to\Omega_K^2$ in Eq. \ref{ode_w} before eliminating
 variables in favor of $\delta v_z$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\upsilon}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\upsilon}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii h q \hat{k} f(\zhat)
      }{1-\hat{\upsilon}^2} \right]W^\prime +
  \hat{\upsilon}^2\left(1+\frac{\hat{k}^2}{1-\hat{\upsilon}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = h^2q f(\hat{z})\Omega_K^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+h^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

\subsection{Stability conditions in the low-frequency limit}
In the low-frequency limit we assume $|\hat{\upsilon}|\ll 1$ and
approximate Eq. \ref{iso_ode0} as 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii h q \hat{k}
    f(\zhat)\right]W^\prime +
  \hat{\upsilon}^2\left(1+\hat{k}^2\right)W. \label{iso_ode1}
\end{align}
\subsubsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  \hat{\upsilon}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat
  =\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii h q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\upsilon}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. there must
be vertical shear. A similar conclusion can be reached in the
equivalent linear problem in global cylindrical disk geometry.

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\upsilon} = \hat{\omega} +
% \ii\hat{\upsilon}$ with real $\hat{\omega}$ and $\hat{\upsilon}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\sigma}^2\right)\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\sigma}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\upsilon}$ and
$\hat{\sigma}=\imag\hat{\upsilon}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\upsilon}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\sigma}|\leq
  \frac{h |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|) < \frac{h |q|}{2}\mathrm{max}(|f|) . 
\end{align}
%Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), 
It
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered,
\begin{align}
  \sigma < \mathrm{max} \left|r\frac{d\Omega}{dz}\right|, 
\end{align}
as expected on physical grounds.

@@

\subsection{Explicit solutions in the thin-disk limit}\label{iso_explicit}
Here we assume a thin disk ($h\ll1$) so that $\ln\rho\simeq
-\zhat^2/2$ and $f(\zhat)\simeq \zhat$. However, we do not assume 
low frequency from the onset. Then Eq. \ref{iso_ode0} becomes 
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \frac{\ii qh \hat{k}}{1-\hat{\upsilon}^2}\right)\hat{z}W^\prime  +
  \hat{\upsilon}^2\left(1+\frac{\hat{k}^2}{1-\hat{\upsilon}^2}\right)W = 0.
\end{align}
We remark that taking the low frequency limit and considering
$\hat{k}^2\gg 1$, Eq. \ref{iso_ode3} becomes equivalent to Eq. 39 in
\citetalias{nelson13} or Eq. 28 in \citetalias{barker15}, although we have taken a
different route.
